---
title: "HELPmiss (Data Verbs) Activity"
author: "YOUR NAME HERE"
output: html_notebook
---


## Set Up:

```{r message=FALSE}
rm(list = ls())     # clean up your R environment

# load packages
)library(tidyverse)  # includes lots of data verbs like `group_by()` and `summarise()`
library(mosaicData # includes the `HELPmiss` data set 


# Load the `HELPmiss` data set into our RStudio environment
data("HELPmiss", package = "mosaicData")


## Helpful links: 

- Look though the DC Textbook for `tidyverse` functions. 
- Check out some of these RStudio cheat sheets:
  - <https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf> 
  - <https://www.rstudio.com/resources/cheatsheets/>
  - <https://s3.amazonaws.com/assets.datacamp.com/blog_assets/Tidyverse+Cheat+Sheet.pdf>
  



## Task 1: Data Description

*Write several sentences (or a bullet list) describing the HELP Study and the resulting `HELPmiss` data.  Your description should investigate basic data provenance (e.g., Who, What, When, Where, Why, How), explain the setting for the data, specify what each case represents in the data, and remark on inclusion/exclusion criteria.* 
Who: The data was collected by the research team involved in the HELP Study.
What: The dataset contains information on patients' demographics, medical diagnoses, treatment interventions, and clinical outcomes.
When: The data collection occurred during the duration of the HELP Study.
Where: The data was collected in a hospital or healthcare setting, where the study was conducted.
Why: The purpose of the study and the resulting dataset is to understand the effectiveness of interventions in improving patient health outcomes.
How: The data was collected through various methods, such as medical records, patient interviews, and observations.
The data represents patients who were admitted to the hospital where the HELP Study was conducted. It provides insights into the medical conditions, treatments, and outcomes of these patients during their hospital stay.
Case representation:
Each case in the HELPmiss data represents an individual patient who participated in the HELP Study. The data provides information on their specific characteristics, medical history, interventions received, and subsequent clinical outcomes.
inclusion/exclusion criteria:
The dataset may have specific inclusion and exclusion criteria that determine which patients were eligible to participate in the HELP Study and, consequently, included in the HELPmiss data. These criteria may include factors such as age, specific medical conditions, and willingness to participate in the study. The exact criteria can be further investigated in the documentation or research publications related to the HELP Study.

- investigate by searching R help documentation
- there's a research paper cited where additional detail is provided



## Task 2: Basic Summaries 

*Produce one or more R expressions involving `summarize()` and `HELPmiss` to address each of the following prompts.*

1. number of people (cases) in `HELPmiss` study

```{r}
summarize(HELPmiss, n = n())


```

2. total number of times in the past 6 months entered a detox program (measured at baseline) for all the people in `HELPmiss`. 

```{r}
summarize(HELPmiss, total_detox_entries = sum(detox_baseline))


```


3. mean time (in days) to first use of any substance post-detox for all the people in `HELPmiss`


```{r}
summarize(HELPmiss, mean_time_post_detox = mean(time_to_first_use))


```



## Task 3: Group Summaries  

*Repeat task 2 above, but add code chunks to calculate result group-by-group according to each prompt.  Be sure to show all R code and write a sentence or two about what you observe in the results. Remember, you can add multiple statistics inside `summary`.* 


- males versus females
- homeless or not
- substance
- break down the homeless versus housed further, by sex
- homeless versus housed broken down by substance

### males versus females
```{r}
by_gender <- group_by(HELPmiss, gender)
summary_gender <- summarize(by_gender, 
                            mean_age = mean(age),
                            median_days_detox = median(detox_baseline),
                            max_time_post_detox = max(time_to_first_use))



```

### homeless or not
```{r}
by_homeless_sex <- group_by(HELPmiss, homeless, gender)
summary_homeless_sex <- summarize(by_homeless_sex, 
                                  mean_age = mean(age),
                                  median_days_detox = median(detox_baseline),
                                  max_time_post_detox = max(time_to_first_use))

```


### substance
```{r}
by_substance <- group_by(HELPmiss, substance)
summary_substance <- summarize(by_substance, 
                               mean_age = mean(age),
                               median_days_detox = median(detox_baseline),
                               max_time_post_detox = max(time_to_first_use))


```

### homeless versus housed broken down by sex
```{r}
by_homeless_sex <- group_by(HELPmiss, homeless, gender)
summary_homeless_sex <- summarize(by_homeless_sex, 
                                  mean_age = mean(age),
                                  median_days_detox = median(detox_baseline),
                                  max_time_post_detox = max(time_to_first_use))

```

### homeless versus housed broken down by substance
```{r}
by_homeless_substance <- group_by(HELPmiss, homeless, substance)
summary_homeless_substance <- summarize(by_homeless_substance, 
                                        mean_age = mean(age),
                                        median_days_detox = median(detox_baseline),
                                        max_time_post_detox = max(time_to_first_use))

```



## Task 4: Data Visualization & Observations

*Include one or more interesting plots from this data set involving at least 3 variables per plot. Write a few sentences to explain the story that your plot tells about these data.  You can expand on the relationships that you studied in Task 2, or you can explore a different group of variables in `HELPmiss` that show something interesting. Remember to use the interactive commands in the console, generate the R commands that will reproduce your plot, and then paste the R commands into an R chunk in the RMarkdown file.*  


```{r}
library(ggplot2)

ggplot(HELPmiss, aes(x = age, y = time_to_first_use, color = substance)) +
  geom_point() +
  labs(x = "Age", y = "Time to First Use Post-Detox", color = "Substance") +
  theme_minimal()



